---
title: "Clustering and PCA"
output:
  pdf_document: default
  html_document:
    df_print: paged
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Dropbox/Teaching/")
```

```{r, echo = T, message = FALSE, results='hide', warning=FALSE}
library(ISLR)
library(factoextra)
library(gridExtra)
library(corrplot)
library(RColorBrewer) 
library(gplots)
```

The dataset we use contains data on 166 first generation Pokemon, including their names and basic stats: HP, Attack, Defense, Special Attack, Special Defense, and Speed. The data is from Kaggle (https://www.kaggle.com/abcsds/pokemon). We will apply unsupervised learning methods on this data. The list of Pokemon can be found at (https://pokemondb.net/pokedex/national).


```{r}
dat <- read.csv("./SlidesCode_Yifei/Pokemon.csv")
dat1 <- dat[,2:7]
dat1 <- scale(dat1)
rownames(dat1) <- dat[,1]
```

# K means clustering

Partitioning methods such as k-means clustering require the users to specify the number of clusters to be generated. The function `fviz_nbclust()` determines and visualizes the optimal number of clusters using different methods: within cluster sums of squares, average silhouette and gap statistics. We use average silhouette, and the greater the silhouette value the better.

```{r, fig.height=3.5}
fviz_nbclust(dat1,
             FUNcluster = kmeans,
             method = "silhouette")
set.seed(1)
km <- kmeans(dat1, centers = 2, nstart = 20)
```

The function `fviz_cluster()` provides ggplot2-based visualization of partitioning methods including K means. Observations are represented by points in the plot, using principal components if $p > 2$. An ellipse is drawn around each cluster.

```{r}
km_vis <- fviz_cluster(list(data = dat1, cluster = km$cluster), 
                       ellipse.type = "convex", 
                       geom = c("point","text"),
                       labelsize = 5, 
                       palette = "Dark2") + labs(title = "K-means") 

km_vis
```

# Hierarchical clustering

We can also apply hierarchical clustering on this data. Here we use the Euclidean distance and different types of linkage.

```{r}
hc.complete <- hclust(dist(dat1), method = "complete")
hc.average <- hclust(dist(dat1), method = "average")
hc.single <- hclust(dist(dat1), method = "single")
hc.centroid <- hclust(dist(dat1), method = "centroid")
```

The function `fviz_dend()` can be applied to visualize the dendrogram.

```{r, fig.width=7}
fviz_dend(hc.complete, k = 4,        
          cex = 0.3, 
          palette = "jco", 
          color_labels_by_k = TRUE,
          rect = TRUE, rect_fill = TRUE, rect_border = "jco",
          labels_track_height = 2.5)

ind4.complete <- cutree(hc.complete, 4)

# Who are in the fourth cluster?
dat[ind4.complete == 4,]
```

To display more details, we show the heatmap of the data.

```{r, fig.width = 12, fig.height=7}
#display.brewer.all(n=NULL, type="all", select=NULL, exact.n=TRUE)
col1 <- colorRampPalette(brewer.pal(9, "GnBu"))(100)
col2 <- colorRampPalette(brewer.pal(3, "Spectral"))(2)

heatmap.2(t(dat1), 
          col = col1, keysize=.8, key.par = list(cex=.5),
          trace = "none", key = TRUE, cexCol = 0.75, 
          labCol = as.character(dat[,1]),
          ColSideColors = col2[as.numeric(dat[,"Legendary"])+1],
          margins = c(10, 10))
```

# PCA

The function `prcomp()` can be used to perform PCA.

```{r, fig.height=3}
pca <- prcomp(dat1)
pca$rotation
pca$sdev

var <- get_pca_var(pca)
corrplot(var$cor)
```

The function `fviz_eig()` plots the eigenvalues/variances against the number of dimensions. 

```{r, fig.height=4}
fviz_eig(pca, addlabels = TRUE)
```

The function `fviz_contrib()` can be used to visualize the contribution of variables from the results of PCA.

```{r}
a <- fviz_contrib(pca, choice = "var", axes = 1)
b <- fviz_contrib(pca, choice = "var", axes = 2)
grid.arrange(a, b, nrow = 2)
```

The function `fviz_pca_biplot()` can be used to obtain the biplot of individuals and variables.

```{r, fig.height=4}
fviz_pca_biplot(pca, axes = c(1,2),
                habillage = ifelse(dat$Legendary==TRUE, "Legendary","Not legendary"),
                label = c("var"),
                addEllipses = TRUE) 

fviz_pca_var(pca, col.var = "steelblue", repel = TRUE)
fviz_pca_ind(pca,
             habillage = ifelse(dat$Legendary==TRUE,"Legendary","Not legendary"),
             label = "none",
             addEllipses = TRUE)
```
